Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-07 3.Thesis/Writeup/2.Manuscript.tex
Feedback Stef

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous across science, but treating it can also be a tedious task. If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this \emph{ad hoc} solution may yield wildly invalid results \citep{buur18}. An alternative is to \emph{impute} (i.e., fill in) the missing values in the incomplete observations. Subsequently, statistical inferences can be performed on the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances \citep{buur18}.

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part \(y_{mis}\) of an incomplete dataset is imputed \(M\) times. The imputed data \(y_{imp}\) is combined with the observed data \(y_{obs}\) to create \(M\) completed datasets. On each completed dataset, an analysis of scientific interest is performed. The quantity of scientific interest (e.g., a regression coefficient) is denoted with \(Q\). Since \(Q\) is estimated on each completed dataset, \(M\) separate \(\hat{Q}\)-values are obtained. These \(M\) values are combined into a single pooled estimate \(\bar{Q}\). The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence valid estimate of the true---but unobserved---inference.

@@

{\centering \includegraphics[width=\linewidth]{../Figures/diagram}

@@

\caption{Scheme of the main steps in multiple imputation (where $M=3$)---from an incomplete dataset, to $M$ multiply imputed datasets, to $M$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
\end{figure}

A popular method to obtain imputations is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}. With MICE, imputed values \(y_{imp}\) are drawn from the posterior predictive distribution of the missing values \(y_{mis}\). The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data?

These remain open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}. Moreover, there is no scientific consensus on how to evaluate convergence of imputation algorithms \citep{taka17}. Some default MICE techniques (e.g., `predictive mean matching') might not yield converged states at all \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully.

Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection. After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm \(\theta\) against the iteration number. Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation. Monitoring convergence by inspecting traceplots of these \(\theta\)s is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} \textbf{{[}Add that convergence of univariate thetas cannot guarantee multivariate convergence?{]}}. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed. First, we define which non-convergence diagnostics will be considered, and evaluate how these diagnostics could be appropriate for iterative imputation applications. In the second part of this paper, we will investigate the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in \texttt{R} \citep{R}. The aim of the simulation is to determine whether unbiased, confidence valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}). Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (\textbf{and on which scalar summary of the state space of the algorithm these diagnostics should be applied}). The results of this study are guidelines for identifying non-convergence in iterative imputation algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets. \textbf{{[}add more societal relevance: what is the consequence of non-convergence? see example of non-convergence in Figure 2 (e.g., biased estimates, invalid inference, etc.){]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) in dataset \(y\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{mis}\). The collection of observed data values in \(y\) is denoted by \(y_{obs}\); the missing part of \(y\) is referred to as \(y_{mis}\). For each datapoint in \(y_{mis}\), we sample \(M \times T\) plausible values, where \(M\) is the number of imputations (\(m = 1, 2, \dots, M\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.

@@

There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}. Without mixing, imputation chains do not intermingle nicely, indicating that the distribution of imputed values differs across imputations. Chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values. This may cause under-estimation of the variance within chains, which results in spurious, invalid inferences. Without stationarity, there is trending within imputation chains. Trending implies that further iterations would yield a different set of imputations. Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.

\textbf{{[}SEGWAY to this section{]}} To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call \(j\). We see two traceplots of the chain means for \(j\)---i.e., the average of the imputed values for \(j\) in each imputation \(y_{imp,m}\), plotted against the iteration number. Each line portrays one imputation. The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. \textbf{In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per \(m\)) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. That means that non-convergence has a impact on the mean of \(j\) in \(y_{imp, m}\). This difference (presumably) translates to bias in the \(M\) completed data estimates \(\hat{\mu}_{j,m}\), and consequently also to the pooled estimate \(\bar{\mu}_j\). Non-convergence thus leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

\textbf{{[}MERGE with section below{]}}. Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are summaries of the univariate state space of the algorithm, while MICE is not only concerned with a sigle column, but the entire multivariate distribution of \(y_{imp}\). Ideally, a multivariate \(\theta\) should be monitored. A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature. Implementing this, however, might be somewhat advanced for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. On these grounds, such a \(\theta\) can be considered insufficient too. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. Such a \(\theta\) would be a model-independent, multivariate scalar summary to monitor, but it has not yet been implemented.

\textbf{{[}MERGE with section above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain \(\theta\). \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\). Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose \(\lambda_1\) as such a scalar summary. We define \(\lambda_1\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_1 \geq \lambda_2 \geq ... \geq \lambda_j\) be the eigenvalues of \(\Sigma\) in each of the \(M\) completed datasets \({y_{obs}, y_{imp}}\). \(\lambda_1\) is measure that summarizes the covariances in the completed datasets. \textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.}

@@

Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for convergence: mixing or stationarity. Non-stationarity within chains may be diagnosed with e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's \citeyearpar{raft91} procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}. Therefore, we use \(\widehat{R}\) and \(AC\) to evaluate mixing and stationarity separately, as recommended by e.g., \citet{cowl96}.

@@

The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}. An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}. \textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to. Namely, rank-normalization, folding, and localization. The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation. We therefore use their definition of the diagnostic.} Let \(M\) be the total number of chains, \(T\) the number of iterations per chain, and \(\theta\) the scalar summary of interest (e.g., chain means). For each chain (\(m = 1, 2, \dots, M\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

\begin{align*}
W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2}, \text { where } s_{m}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. 
\end{align*}

\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain.

\begin{align*}
B&=\frac{T}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t m)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)}. 
\end{align*}

@@

We can interpret \(\widehat{R}\) as potential scale reduction factor since it indicates by how much the variance of \(\theta\) could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. This interpretation assumes that chains are `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\). Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied. The variance between chains is then equivalent to the variance within chains, and \(\widehat{R}\)-values will be close to one. High \(\widehat{R}\)-values thus indicate non-convergence. The conventionally acceptable threshold for convergence was \(\widehat{R} < 1.2\) \citep{gelm92}. More recently, \citet{veht19} proposed a more stringent threshold of \(\widehat{R} < 1.01\).

@@

We can interpret \(AC\)-values as a measure of stationarity. If \(AC\)-values are close to zero, there is no dependence between subsequent samples within imputation chains. Negative \(AC\)-values indicate divergence within imputation chains. \textbf{Subsequent sampled values within each imputation chain are less alike.} Positive \(AC\)-values indicate recurrence. If \(\theta\)-values of subsequent iterations are similar, trending may occur. Negative \(AC\)-values show no threat to the stationarity component of convergence. On the contrary even---negative \(AC\)-values indicate that \(\theta\)-values of subsequent iterations diverge from one another, which may increase the variance of \(\theta\) and speed up convergence. As convergence diagnostic, the interest is therefore in positive \(AC\)-values. The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}.

@@

Upon convergence, \(\widehat{R}=1\) and \(AC=0\), which are unlikely thresholds for MCMC algorithms, because of its convergence to a distribution. Even in the most converged state, the algorithm will show some signs of non-mixing and non-stationarity. Upon \emph{sufficient} convergence, the imputation chains will intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there may be some dependency between subsequent iterations of imputation chains (\(AC > 0\)). In practice, non-convergence is usually diagnosed when \(\widehat{R}\) \textgreater{} 1.2 or 1.1 or even 1.01. And a t-test is performed to assess whether \(AC\) is significantly different from zero.

@@

\caption{Convergence diagnostics applied on the imputation algorithms of Figure 2. $\theta$ = chain mean in $y_{imp, m}$.}\label{fig:diagnostics}
\end{figure}

@@

\textbf{Describe panel C here and why it is not optimal{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above. \(\widehat{R}\) as computed conform \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up. \textbf{\citep[explain the dip in Rhat values at t=2. Namely, because we can only use 2 of the 3 tricks by][ if the number of iterations is very low (t\textless4). That's why the \(\widehat{R}\)s are more similar to the traditional GR.]{veht19}}

@@

The aim of the simulation study is to evaluate the impact of inducing non-convergence in the MICE algorithm on several quantities of scientific interest \(Q\). Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence \textbf{{[}and performance of different thetas{]}}. And finally, we will formulate an informed advice on the requirements to safely assume sufficient convergence in practice. That is, we assume that the algorithm is sufficiently converged when each estimate \(\bar{Q}\) is an unbiased and confidence valid estimate of the corresponding estimand \(Q\). \textbf{{[}and formulate advice on thetas?{]}}

>>>>

Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous across science, but treating it can also be a tedious task. If a dataset contains just one incomplete observation, calculations are not defined, \textbf{hence} statistical models cannot do fitted to the data. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this \emph{ad hoc} solution may yield invalid results \citep{buur18}. An alternative is to apply imputation. With imputation, we `fill in' the missing values in an incomplete dataset. Subsequently, the model of scientific interest can be fitted to the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as `multiple imputation' \citep[MI;][]{rubin76}. MI has proven to be a powerful technique to draw valid inferences from incomplete data under many circumstances \citep{buur18}.

Figure \ref{fig:diagram} provides an overview of the steps involved with MI. The missing part \(y_{\rm mis}\) of an incomplete dataset is imputed \(m\) times. This creates \(m\) sets of imputed data \(\dot y_{\rm imp, \ell}\), where \(\ell = 1, 2, ..., m\). The imputed data is then combined with the observed data \(y_{\rm obs}\) to create \(m\) completed datasets. On each of these datasets the analysis of scientific interest is performed to estimate \(Q\): the quantity of scientific interest (e.g., a regression coefficient). Since \(Q\) is estimated on each completed dataset, \(m\) separate \(\hat{Q}_\ell\)-values are obtained. Finally, the \(m\) \(\hat{Q}_\ell\)-values are combined into a single pooled estimate \(\bar{Q}\). The premise of multiple imputation is that \(\bar{Q}\) is an unbiased and confidence-valid estimate of the true---but unobserved---inference \citep{rubin96}.

@@

{\centering \includegraphics[width=\linewidth]{../Figures/DiagramMI}

@@

\caption{Scheme of the main steps in multiple imputation (where $m=3$)---from an incomplete dataset, to $m$ multiply imputed datasets, to $m$ estimated quantities of scientific interest $\hat{Q}$s, to a single pooled estimate $\bar{Q}$.}\label{fig:diagram}
\end{figure}

A popular method to obtain imputations is to use the `Multiple Imputation by Chained Equations' algorithm, shorthand `MICE'\citep{mice}. With MICE, imputed values \(\dot y_{\rm imp, \ell}\) are drawn from the posterior predictive distribution of the missing values \(y_{\rm mis}\). The algorithm is named after the iterative algorithmic procedure by which imputed values are generated: a multivariate distribution is obtained by iterating over a sequence of univariate imputations. The iterative nature of algorithms like MICE introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained on the completed data?

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}. There is no scientific consensus on how to evaluate convergence of imputation algorithms \citep{taka17}. \textbf{Moreover, the convergence of chained equations algorithms such as MICE under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}.} Therefore, algorithmic convergence should be monitored carefully.

Currently, the recommended practice for evaluating the convergence of iterative imputation algorithms is through visual inspection. After running the imputation algorithm for a certain number of iterations, its convergence is monitiored by plotting a scalar summary of the state space of the algorithm \(\theta\) against the iteration number. Typically, the \(\theta\)s under evaluation are chain means and variances---the average and the variance of the imputed values per imputation. Monitoring convergence by inspecting traceplots of these \(\theta\)s may be undesirable for several reasons on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed \citep[\(\S\) 6.5.2]{buur18} and, 3) there is not an objective measure that quantifies convergence. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred---although not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

In this paper, we study different methods for assessing non-convergence in iterative imputation algorithms. We define several diagnostics and evaluate how these diagnostics could be appropriate for iterative imputation applications. We then adress the impact of inducing non-convergence in iterative imputation algorithms by means of model-based simulation in \texttt{R} \citep{R}. The aim of the simulation is to determine whether unbiased, confidence-valid inferences may be obtained under different levels of non-convergence (\textbf{and at which point convergence may safely be assumed}). Additionally, we will assess how well different non-convergence diagnostics perform in identifying signs of non-convergence (\textbf{and on which scalar summary of the state space of the algorithm these diagnostics should be applied}). We translate the results of the study into guidelines for applied researchers. \textbf{{[}add more societal relevance: what is the consequence of non-convergence? see example of non-convergence in Figure 2 (e.g., biased estimates, invalid inference, etc.){]}} For reasons of brevity, we only focus on the iterative imputation algorithm implemented in the popular \texttt{mice} package \citep{mice} in \texttt{R} \citep{R}.

@@

Let \(y\) denote an \(n \times k\) matrix containing the data values on \(k\) variables for all \(n\) units in a sample. The data value of unit \(i\) (\(i = 1, 2, \dots, n\)) on variable \(j\) (\(j = 1, 2, \dots, k\)) may be either observed or missing. The number of units \(i\) in dataset \(y\) with at least one missing value, divided by the total number of units \(n\), is called the missingness proportion \(p_{\rm mis}\). The collection of observed data values in \(y\) is denoted by \(y_{\rm obs}\); the missing part of \(y\) is referred to as \(y_{\rm mis}\). For each datapoint in \(y_{\rm mis}\), we sample \(m \times T\) plausible values, where \(m\) is the number of imputations (\(\ell = 1, 2, \dots, m\)) and \(T\) is the number of iterations (\(t = 1, 2, \dots, T\)) in the imputation algorithm. The collection of samples between the initial value (at \(t=1\)) and the final imputed value (at \(t=T\)) will be referred to as an `imputation chain'.

@@

There are two requirements for convergence of iterative algorithms: mixing and stationarity \citep{gelm13}. Without mixing, imputation chains do not intermingle nicely, indicating that the distribution of imputed values differs across imputations. Chains may be `stuck' at a local optimum, instead of sampling imputed values from the entire predictive posterior distribution of the missing values. This may cause under-estimation of the variance within chains, which results in spurious, invalid inferences. Without stationarity, there is trending within imputation chains. Trending implies that further iterations would yield a systematically lower or higher set of imputations. Iterative imputation algorithms that have not yet reached stationarity, may thus yield biased estimates.

\textbf{{[}SEGWAY to this section{]}} To illustrate what non-convergence looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays a subset of the example: a variable that we will call \(j\). We see two traceplots of the chain means for \(j\)---i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\), plotted against the iteration number. Each line portrays one imputation. The plot on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. \textbf{In the typical convergence situation, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence plot, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. That means that non-convergence has a impact on the mean of \(j\) in \(\dot y_{\rm imp, \ell}\). This difference (presumably) translates to bias in the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently also to the pooled estimate \(\bar{\mu}_j\). Non-convergence thus leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

\textbf{{[}MERGE with section below{]}}. Inspecting non-convergence through traceplots of chain means and chain variances may be insufficient because they are univariate summaries of the state space of the algorithm, while MICE is not only concerned with a single column, but the entire multivariate distribution of \(y_{\rm imp}\). Ideally, a multivariate \(\theta\) should be monitored. A suggestion by \citet{buur18} for a multivariate \(\theta\) to monitor is the quantity of scientific interest \(Q\) (e.g., a regression coefficient), which usually is multivariate in nature. Implementing this, however, might be somewhat too technical for empirical researchers. Moreover, this scalar summary is not model-independent, i.e., it only applies to one model of scientific interest. Yet, one of the advantages of MI is that the missing data problem and scientific problem are solved independently. On these grounds, such a \(\theta\) can be considered insufficient too. Van Buuren \citeyearpar[\(\S\) 4.5.2]{buur18} also proposed multivariable evaluation of the MICE algorithm through eigenvalue decomposition, building on the work of \citet{mack03}. Eigenvalues of a variance-covariance matrix are a measure of the data's total covariance. Such a \(\theta\) would be a model-independent, multivariate scalar summary to monitor, but it is not implemented in \texttt{mice}.

\textbf{{[}MERGE with section above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain \(\theta\). \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration. The default scalar summaries in \texttt{mice} are chain means and chain variances. Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\). Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose \(\lambda_{1, \ell}\) as such a scalar summary. We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. \(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets. \textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.}

@@

Non-convergence diagnostics for MCMC algorithms typically identify problems in either one of the two requirements for convergence: mixing or stationarity. Non-stationarity within chains may be diagnosed with e.g., autocorrelation \citep[\(AC\);][]{scha97, gelm13}, numeric standard error \citep[`MC error';][]{gewe92}, or Raftery and Lewis's \citeyearpar{raft91} procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor \(\widehat{R}\) \citep[`Gelman-Rubin statistic';][]{gelm92}. With a recently proposed adaptation, \(\widehat{R}\) might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated \citep{veht19}. Therefore, we use \(\widehat{R}\) \textbf{{[}WHICH? Explain that we'll use both{]}} and \(AC\) to evaluate mixing and stationarity separately, as recommended by e.g., \citet{cowl96}.

@@

The potential scale reduction factor \(\widehat{R}\) was coined by \citet{gelm92}. An updated version of \(\widehat{R}\) has been proposed by Vehtari et al.~\citeyearpar[p.~5]{veht19}. \textbf{This version is better suited to detect non-convergence in the tails of distributions, by using three transformations on the scalar summary that it is applied to. Namely, rank-normalization, folding, and localization. The Vehtari et al.~version of \(\widehat{R}\) may be suitable for iterative imputation. We therefore use their definition of the diagnostic.} Let \(m\) be the total number of chains, \(T\) the number of iterations per chain (\(T\geq2\)), and \(\theta\) the scalar summary of interest (e.g., chain means). For each chain (\(\ell = 1, 2, \dots, m\)), we estimate the variance of \(\theta\), and average these to obtain within-chain variance \(W\).

\begin{align*}
W&=\frac{1}{m} \sum_{\ell=1}^{m} s_{j}^{2}, \text { where } s_{\ell}^{2}=\frac{1}{T-1} \sum_{t=1}^{T}\left(\theta^{(t \ell)}-\bar{\theta}^{(\cdot \ell)}\right)^{2}. 
\end{align*}

\noindent We then estimate between-chain variance \(B\) as the variance of the collection of average \(\theta\) per chain. \textbf{{[}Note that this is not the usual B in MI{]}}

\begin{align*}
B&=\frac{T}{m-1} \sum_{\ell=1}^{m}\left(\bar{\theta}^{(\cdot \ell)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot \ell)}=\frac{1}{T} \sum_{t=1}^{T} \theta^{(t \ell)} \text{, } \bar{\theta}^{(\cdot \cdot)}=\frac{1}{m} \sum_{\ell=1}^{m} \bar{\theta}^{(\cdot \ell)}. 
\end{align*}

@@

We can interpret \(\widehat{R}\) as potential scale reduction factor since it indicates by how much the variance of \(\theta\) could be shrunken down if an infinite number of iterations per chain would be run \citep{gelm92}. This interpretation assumes that chains are `over-dispersed' at \(t=1\), and reach convergence as \(T \to \infty\). Over-dispersion implies that the initial values of the chains are `far away' from the target distribution and each other. When the sampled values in each chain are independent of the chain's initial value, the mixing component of convergence is satisfied. The variance between chains is then equivalent to the variance within chains, and \(\widehat{R}\)-values will be close to one. High \(\widehat{R}\)-values thus indicate non-convergence.

@@

We can interpret \(AC\)-values as a measure of stationarity. \(AC\)-values close to zero indicate independence between subsequent \(\theta\)s in imputation chains. Negative \(AC\)-values show no threat to the stationarity component of convergence. On the contrary even---negative \(AC\)-values indicate that \(\theta\)-values of subsequent iterations diverge from one another, which may increase the variance of \(\theta\) and speed up convergence. As convergence diagnostic, the interest is therefore in positive \(AC\)-values. Positive \(AC\)-values occur when \(\theta\)-values of subsequent iterations are similar. When high \(\theta\)-values are followed by high \(\theta\)-values, and low \(\theta\)-values are followed by low \(\theta\)-values, we speak of recurrence. Recurrence within imputation chains indicates trending. Ergo, non-stationarity may be diagnosed when \(AC\) values are larger than zero.

@@

It is unlikely that an MCMC algorithm will achieve the ideal values of \(\widehat{R}=1\) and \(AC=0\). Because of its convergence to a distribution, the algorithm will show some signs of non-mixing and non-stationarity even in the most converged state. Upon approximate convergence, the imputation chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm (\(\widehat{R} > 1\)), and there is some dependency between subsequent iterations of imputation chains (\(AC > 0\)). In practice, non-convergence is therefore diagnosed when non-convergence diagnostics exeed a certain threshold. The conventional threshold to diagnose non-mixing is \(\widehat{R} > 1.2\) \citep{gelm92}. Another generally accepted threshold is \(\widehat{R} > 1.1\) \citep{gelm13}. And more recently, \citet{veht19} proposed an even more stringent threshold of \(\widehat{R} > 1.01\). The magnitude of \(AC\)-values may be evaluated statistically, using a Wald test with \(AC=0\) as null hypothesis \citep{box15}. \(AC\)-values that are significantly higher than zero indicate non-stationarity. \textbf{{[}think about one-sided test{]}}

@@

\caption{Convergence diagnostics applied on the imputation algorithms of Figure 2. $\theta$ = chain mean in $ \dot y_{\rm imp, \ell} $.}\label{fig:diagnostics}
\end{figure}

@@

\textbf{Describe panel C here and why it is not optimal: NB. In panel C: De grotere variantie overall zorgt er kennelijk voor dat \(\widehat{R}\) groeit. De \(\widehat{R}\) geeft de foutieve boodschap dat convergentie steeds slechter gaat. Verklaring: de overdispersie assumptie klopt niet als er trend is.{]}} The \(\widehat{R}\)-values in panel D do meet the requirements specified above. \(\widehat{R}\) as computed conform \citet{veht19} does indeed indicate less signs of non-convergence as the number of iterations goes up. \textbf{\citep[explain the dip in Rhat values at t=2. Namely, because we can only use 2 of the 3 tricks by][ if the number of iterations is very low (t\textless4). That's why the \(\widehat{R}\)s are more similar to the traditional GR.]{veht19}}

@@

The aim of the simulation study is to evaluate the \textbf{impact of inducing non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} in the MICE algorithm on several quantities of scientific interest \(Q\). Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence \textbf{{[}and performance of different thetas{]}}. And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice. That is, we conclude that convergence is sufficient when each estimate \(\bar{Q}\) is an unbiased and confidence-valid estimate of the corresponding estimand \(Q\). \textbf{{[}and formulate advice on thetas?{]}}
